---
title: "Examples"
author: "Konrad Krämer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to use *ast2ast*

### R code is translated

```{r}
add_two <- function(a) {
  return(a + 2)
}  
pointer_to_f_cpp <- ast2ast::translate(add_two)
```

### Defining C++ Code which uses the translated R Code

```{Rcpp}
// [[Rcpp::depends(ast2ast)]]
// [[Rcpp::depends(RcppArmadillo)]]
#include "etr.hpp"

// [[Rcpp::plugins("cpp17")]]

typedef sexp (*fp)(sexp a);

// [[Rcpp::export]]
void call_fct(Rcpp::XPtr<fp> inp) {
  fp f = *inp;
  sexp a = coca(1, 2, 3);
  print(a);
  a = f(a);
  print("a is now:");
  print(a);
}
```

### Calling the final function

```{r}
#Rcpp::sourceCpp("cpp_code.cpp") # if not run in Rmarkdown
call_fct(pointer_to_f_cpp)
```







## Introduction to additional functions

### R code is translated

```{r}
add_two <- function(a, b) {
  d <- matrix(3, 3, 4)
  c <- a + b - d/2 + sin(3.14159)
  return(c)
}  
pointer_to_f_cpp <- ast2ast::translate(add_two)
```

### Defining C++ Code which uses the translated R Code

```{Rcpp}
// [[Rcpp::depends(ast2ast)]]
// [[Rcpp::depends(RcppArmadillo)]]
#include "etr.hpp"

// [[Rcpp::plugins("cpp17")]]

typedef sexp (*fp)(sexp a, sexp b);

// [[Rcpp::export]]
void call_fct(Rcpp::XPtr<fp> inp) {
  fp f = *inp;
  sexp a = matrix(3.14, 3, 4);
  sexp b = matrix(5, 3, 4);
  
  print(a);
  print(b);
  
  sexp c = f(a, b);
  print();
  print(c);
}
```

### Calling the final function

```{r}
#Rcpp::sourceCpp("cpp_code.cpp") # if not run in Rmarkdown
call_fct(pointer_to_f_cpp)
```













## Using *ast2ast* in combination with *r2sundials*

### R code is translated

```{r}
ode <- function(y, ydot) {
  nu <- 2
  a <- 1
  ydot <- -nu*(y - a)
} 

pointer_to_ode <- ast2ast::translate(ode, reference = TRUE, verbose = FALSE)
```

### Defining C++ Code which uses the translated R Code

```{Rcpp}

// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(rmumps)]]
// [[Rcpp::depends(r2sundials)]]
// [[Rcpp::depends(ast2ast)]]
#include "etr.hpp"
#include "RcppArmadillo.h"

#include "r2sundials.h"

using namespace arma;

// [[Rcpp::plugins("cpp17")]]

typedef int (*fp)(double t, const vec &y, vec &ydot, RObject &param, NumericVector &psens);

typedef void (*user_fct)(sexp& y_, sexp& ydot_);
user_fct Fct;

int rhs_exp(double t, const vec &y, vec &ydot, RObject &param, NumericVector &psens) {
  NumericVector p(param);
  double nu = 2;
  double a = 1;
  ydot[0] = -nu*(y[0]-a);
  return(CV_SUCCESS);
}


int rhs_exp_wrapper(double t,const vec &y, vec &ydot, RObject &param, NumericVector &psens) {
  NumericVector p(param);
  const int size = y.size(); 
  sexp ydot_(size, ydot.memptr(), 2);
  // bad --> but otherwise r2sundials cannot be used
  double* ptr = const_cast<double*>(y.memptr());
  sexp y_(size, ptr, 2);
  // user code starts
  Fct(y_, ydot_);
  // user code ends
  
  return(CV_SUCCESS);
}



// [[Rcpp::export]]
Rcpp::NumericVector test_ode() {

  Rcpp::XPtr<fp> ptr = Rcpp::XPtr<fp>(new fp(&rhs_exp));

  Rcpp::Environment pkg = Rcpp::Environment::namespace_env("r2sundials");
  Rcpp::Function f = pkg["r2cvodes"];


  NumericVector ti{0, 1, 2, 3};
  NumericVector y0{0};

  Rcpp::NumericVector output = f(y0, ti, ptr, ti);

  return output;
}




// [[Rcpp::export]]
Rcpp::NumericVector test_ode_etr(Rcpp::XPtr<user_fct> inp) {
  Fct = *inp;  
  Rcpp::XPtr<fp> ptr = Rcpp::XPtr<fp>(new fp(&rhs_exp_wrapper));
  
  Rcpp::Environment pkg = Rcpp::Environment::namespace_env("r2sundials");
  Rcpp::Function f = pkg["r2cvodes"];


  NumericVector ti{0, 1, 2, 3};
  NumericVector y0{0};

  Rcpp::NumericVector output = f(y0, ti, ptr, ti);

  return output;
}
 
```

### Calling the final function

```{r}
# R version
ti <- c(0, 1, 2, 3) 
p <- c(nu = 2, a = 1)
y0 <- 0
frhs <- function(t, y, p, psens) -p["nu"]*(y-p["a"]) 
library(r2sundials)

res_pointer <- test_ode()
res_ast2ast <- test_ode_etr(pointer_to_ode)
res_R <- r2cvodes(y0, ti, frhs, param=p)

print("Do we get the same results?")
if(diff(range(res_pointer - res_ast2ast)  < 1.e-14) == 0) print("yes")
if(diff(range(res_R - res_ast2ast)  < 1.e-14) == 0) print("yes")

# Benchmark
microbenchmark::microbenchmark(test_ode(), test_ode_etr(pointer_to_ode), r2cvodes(y0, ti, frhs, param=p))
```

