---
title: "Information for package authors"
author: "Konrad Krämer"
output: html_document

vignette: >
  %\VignetteIndexEntry{Information for package authors}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, cache = FALSE)
```


<style>
body {
text-align: justify}
</style>

```{css echo=FALSE}
/* Define a margin before h2 element */
h2  {
  margin-right: 6em;
}

/* Define a margin after every first p elements */
p:first-of-type {
  margin-bottom: 3em;
}
``` 

## Guidance for Package Authors

This section of the documentation describes how the external pointers to C++ function that are produced by *ast2ast* can be used in other packages.
This information is intended for package authors who want to use the translated functions within their own code.
Additionally, this section endeavors to present pertinent implementation details, providing valuable insights into the inner workings of the process.

## Naming rationalization ast2ast

The nomenclature of the package, ast2ast, is rooted in the abbreviation *ast*, signifying abstract syntax tree. 
Originally I planned to convert the abstract syntax tree of R to the C++ tree as the literature recommended for transpilers.
However, through iterative refinement, a more optimal methodology emerged.
The development incorporated an Expression Template Library in C++ known as ETR, meticulously crafted to mimic R.
Thus, R code is translated into ETR code, which is subsequently compiled.
The original ETR library is accessible at https://github.com/Konrad1991/ETR. 
It's imperative to note that the version integrated into ast2ast has undergone substantial enhancements, amplifying its efficacy and adaptability.

## Comparison of R and ETR code

Displayed below is a basic bubble sort function implemented in R on the left, juxtaposed with its ETR counterpart on the right.
It is obvious that the two code snippets are quite similar. 
Remarkably, the overall structure of the R code remains unaltered. 
Instead, the substitution of individual functions with their ETR equivalents forms the crux of the transformation. \
In the C++ code, all functions are located in the *etr* namespace.
Certain functions share identical names in both R and C++, such as the *length* function.
To mitigate potential conflicts, these calls are modified to explicitly reference the etr namespace, 
resulting in expressions like *etr::length*. Other functions as for example *`:`* and *`[`* cannot be defined in C++ (at least not in the way they are used in R)
therefore they are replaced by functions with new names e.g. *etr::colon* and *etr::subset*. \
Additionally, C++ necessitates explicit declaration of variable types. 
Within this example for all variables the type *Array<Double, Buffer<Double>>* is used. A detailed explanation is deferred to the subsequent section.

```{r, eval = FALSE, echo = TRUE, attr.source='.numberLines'}
bubbleSort <- function(a) {
  size <- length(a)
  for (i in 1:size) {
    for (j in 1:(size - 1)) {
      if (a[j] > a[j + 1]) {
        temp <- a[j]
        a[j] <- a[j + 1]
        a[j + 1] <- temp
      }
    }
  }
  return(a)
}
```

```{Rcpp, eval = FALSE, echo = TRUE, attr.source='.numberLines'}
// [[Rcpp::depends(ast2ast)]]
// [[Rcpp::plugins(cpp2a)]]
#include "etr.hpp"
// [[Rcpp::export]]
SEXP bubbleSort(SEXP aSEXP) {
etr::Aetr::Double, etr::Buffer<etr::Double>> a = aSEXP;
  int size;
etr::Array<etr::Double, etr::Buffer<etr::Double>> temp;
  size = etr::length(a);
  for(const auto& i : etr::colon(1, size)) {
   for(const auto& j : etr::colon(1, (size - 1.0))) {
    if (etr::subset(a, j) > etr::subset(a, j + 1.0)) {
       temp = etr::subset(a, j);
       etr::subset(a, j) = etr::subset(a, j + 1.0);
       etr::subset(a, j + 1.0) = temp;
    };
   }
  }
  return(etr::Cast(a));
}
```


### The XPtr interface 

The XPtr interfaces creates an external pointer which can be used in other C++ programs. 

#### First, one creates the external pointer of the R function. 

```{r, eval = TRUE, echo = TRUE}
f <- function(a, b) {
  c <- a + b
  return(c)
}
f_args <- function(a, b) {
  a |> type(vec(double)) |> ref()
  b |> type(vec(double)) |> ref()
}
fcpp <- ast2ast::translate(f, f_args, output = "XPtr")
```

#### Afterwards the external pointer is used in an R package

```{Rcpp, eval = TRUE, echo = TRUE}
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(ast2ast)]]
// [[Rcpp::plugins(cpp2a)]]
#include "etr.hpp"

typedef etr::Array<etr::Double, etr::Buffer<etr::Double>> (*FP)(etr::Array<etr::Double, etr::Buffer<etr::Double>>& a, etr::Array<etr::Double, etr::Buffer<etr::Double>>& b);

// [[Rcpp::export]]
void call_xptr(Rcpp::XPtr<FP> ep) {
  FP f = *ep;
  etr::Array<etr::Double, etr::Buffer<etr::Double>> a;
  etr::Array<etr::Double, etr::Buffer<etr::Double>> b;
  etr::Array<etr::Double, etr::Buffer<etr::Double>> c;
a = etr::c(etr::Double(1), etr::Double(2), etr::Double(3));
b = etr::c(etr::Double(4), etr::Double(5), etr::Double(6));
  c = f(a, b);
  etr::print(c);
}
```
#### call the function from the R package


```{r, eval = TRUE, echo = TRUE}
call_xptr(fcpp)
```

